A unified pseudo-$C_\ell$ framework

### 2018-07-31 doc/Paper/main.tex
show effects of deprojection bias
Added 4096 validation, corrected typos

    The paper is structured as follows: Section \ref{sec:maths} introduces the pseudo-$C_\ell$ estimator in the curved sky, including mode deprojection and $E/B$ purification. This is then generalized in the flat-sky approximation. In Section \ref{sec:structure} we briefly describe the implementation of these methods on \nmt, as well as its structure and usage. Section \ref{sec:validation} presents the stress tests used to validate the code. We envisage this code to be useful in analysis of both the Cosmic Microwave Background data and in the tomographic large-scale structure from photometric data and this section illustrates not just its validity but also applicatiblity in typical user-case scenarios in both fields. We conclude in Section \ref{sec:discussion}. The appendices provide further details about the mathematical background used throughout the paper, and present some additional validation tests.

@@

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \va^\dagger \mC^{-1}
          \left(\frac{\partial \mC}{\partial B_i} \right)
          \mC^{-1} \va - \rm{Tr}\left[\left(\frac{\partial \mC}{\partial B_i}
          \right) {\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix,the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}
          \mC = \sum_i \left(\frac{\partial \mC}{\partial B_i}\right) + {\sf N},
        \end{equation}
        and the last term correspond to bias of convolved estimates by the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices $\partial {\sf C}/\partial B_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particuarly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

      Two further points must be addressed before moving on. First, the computation of the deprojection bias requires an estimate of the true power spectrum ${\sf C}^{ab}_\ell$ (see Eqs. \ref{eq:dp.g.aux} and \ref{eq:dp.f.aux}). In practice, this can be done through an iterative approach, starting from an initial guess of ${\sf C}^{ab}_\ell$ and replacing it by the estimated power spectrum in the previous iteration. A sufficiently good guess will likely be enough to remove the bias in a single iteration. Alternatively, the ensemble average of Eq. \ref{eq:dpbias0} can be seen as a convolution of ${\sf C}^{ab}_\ell$, which could be treated by computing the convolution matrix as a modified version of the standard mode-coupling matrix that accounts for mode deprojection. Estimating this convolution matrix in this case becomes significantly more computationally intensive, and therefore an iterative approach is preferred. \as{I think this is not quite true. This latter way of looking at it, in effect replaces the true power spectrum with its observed quantity, which is fine in practice, but conceptually not quite right, for the same raeason e.g. error on power spectrum should be based on real rather than measured power as illustrated by low CMB quadrupole. I would just remove text after ``Alternatively...''} \dam{we can do this, but I'd like to understand what the problem with this is. Since you can understand a mask as a set of deprojected contaminants (one per masked pixel), the pseudo-$C_\ell$ mode-coupling matrix is basically a flavour of the equations above for one particular set of templates. In this sense then writing the equation above as a matrix-vector product and inverting that matrix (after binning etc.) seems like the pseudo-$C_\ell$ way to go. Are you saying that you'll get more optimal results by putting in a guess for the power spectrum because that's closer to what a QML estimator would do?}

@@

    The code is written in C and wrapped into {\tt python} to facilitate its use and its combination with other software libraries for astronomy. \nmt~itself is based on a number of these libraries:
    \begin{itemize}
      \item {\tt HEALPix} \citep{2005ApJ...622..759G} is currently the only pixelization scheme supported for full-sky calculations. Through {\tt HEALPix}, \nmt~also makes use of {\tt cfitsio}~\citep{1999ASPC..172..487P}.
      \item The basic full-sky operation carried out by \nmt~is the spherical harmonic transform. For this, the code makes use of {\tt libsharp}~\citep{2013A&A...554A.112R}.
      \item In flat sky, SHTs are replaced by 2D DFTs, for which \nmt~uses the {\tt FFTW} library~\citep{FFTW05}.
    \end{itemize}

@@

    \paragraph*{NmtField.} This class stores all the necessary information about one individual observd spin-0 or spin-2 field. A {\tt NmtField} object is defined by a sky mask or weights map (e.g. $v$ in Eq. \ref{eq:aobs}), 1 or 2 sky maps (depending on the spin), corresponding to an observation of the field on the sky (${\bf a}^{\rm obs}$ in Eq. \ref{eq:aobs}), a set of contaminant templates (${\bf f}_i$ in Eq. \ref{eq:aobs}), and the choice to purify or not the $E$ or $B$ mode component of the field. Once initialized, this class carries out the following operations:
    \begin{itemize}
      \item SHT or DFT of the field (Eqs. \ref{eq:sht1} or \ref{eq:dft}).
      \item Contaminant deprojection (Eq. \ref{eq:clean}).
      \item $E$-mode and/or $B$-mode purification (Eq. \ref{eq:pureb}).
    \end{itemize}

@@

        Large-scale structure data are often characterized by inhomogeneous sky coverages and complex mask structures, associated with the presence of bright stars and observational artifacts. They are also affected by many sky contaminants, associated to Galactic sources (dust absorption, star density) and observing conditions (seeing, airmass etc.)~\citet{2016ApJS..226...24L}. The typically flatter power spectra for both galaxy clustering and lensing (e.g. when compared with the damping tail of CMB temperature fluctuations), makes the pseudo-$C_\ell$ approach ideal for this kind of analysis~\citet{2013MNRAS.435.1857L,2016ApJS..226...24L}. This science case therefore allows us to validate several aspects of the code simultaneously:
        \begin{itemize}
          \item The unbiased reconstruction of the 6 possible correlations between spin-0 and spin-2 fields.
          \item The inclusion of inhomogeneous noise and the correction of the associated noise bias.
          \item The effects of deprojection for several contaminant templates in different-spin fields.
          \item The code's ability to handle complex masks and high-resolution pixelization.
        \end{itemize}

@@

    \subsection{Full-sky large-scale structure validation}\label{ssec:validation.lss_full}
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_lss_sph}
        \includegraphics[width=0.99\columnwidth]{val_covar_lss_sph}
        \caption{{\sl Left panel:} average over all simulations in our full-sky LSS validation suite of the six power spectra between $\delta$, $\gamma_E$ and $\gamma_B$ (solid lines). The analytical prediction for the input signal power spectra are shown as dashed lines, and lie mostly below the solid lines (for the non-zero cases $\delta\times\delta$, $\delta\times\gamma_E$ and $\gamma_E\times\gamma_E$). The null power spectra ($\delta\times\gamma_B$, $\gamma_E\times\gamma_B$ and $\gamma_B\times\gamma_B$) are compatible with zero. The bottom subplot shows the mean residuals with respect to the input power spectra normalized by the 1$\sigma$ error over the mean (i.e. the scatter over simulations divided by $\sqrt{N_{\rm sim}}$). The residuals are scattered around zero and their fluctuations are well within those expected under the assumption of Gaussian statistics given 1000 independent realizations (see Fig. \ref{fig:lss_val_chi2}). {\sl Right panel:} correlation matrix ($C_{ij}/\sqrt{C_{ii}C_{jj}}$) of all measured power spectra estimated from 1000 realizations.} \label{fig:lss_val_cls}
      \end{figure*}
      \begin{figure*}
        \centering
        \includegraphics[width=0.80\textwidth]{val_chi2_lss_sph}
        \caption{Distribution of the $\chi^2$ values found for each of the 1000 simulations in our full-sky LSS validation suite (blue histograms). Results are shown for all possible cross-correlations, labeled in the top right corner of each subplot. The simulated distributions are compared with the expected $\chi^2$ distribution (solid lines) assuming Gaussian statistics for $N_{\rm dof}=XXX$ degrees of freedom (denoted by the dashed vertical lines). The dot-dashed vertical lines mark the $\chi^2$ value found for the mean residual with respect to the input power spectrum (see lower half of the left panel in Fig. \ref{fig:lss_val_cls}). The associated PTE values are $\gtrsim0.1$, and thus no statistically significant bias is found by our validation suite.} \label{fig:lss_val_chi2}
      \end{figure*}
      We generate 1000 realizations of contaminated maps of the galaxy overdensity $\delta$ and cosmic shear $\gamma$ using the models described in Section \ref{sssec:validation.suite.lss} with resolution $N_{\rm side}=1024$. In our fiducial run, we included all contaminants, and used a non-binary mask that traces the shot-noise fluctuations due to the presence of stars and depth variations. For each simulation, we use \nmt~to compute all six possible cross-correlations between the $\delta$ and the $E$ and $B$ modes of $\gamma$ (which we will label $\gamma_{E.B}$), incuding map-level contaminant deprojection and removal of the noise and deprojection bias in the power spectrum. We use band-powers defined by averaging over groups of $\Delta\ell=XX$ multipoles.

@@

      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_cmb_sph}
        \includegraphics[width=0.99\columnwidth]{val_cmb_bpw}
        \caption{{\sl Left:} average over all simulations in our full-sky CMB validation suite of the three spin-2 power spectra ($EE$, $EB$ and $BB$, solid lines). The analytical prediction for the input signal power spectra are shown as dashed lines, and line mostly below the solid lines for the non-zero cases ($EE$ and $BB$). The null power spectrum $EB$ is compatible with zero. The bottom panel of this figure shows the mean residuals with respect to the input power spectra normalized by the $1\sigma$ error on the mean (i.e. the scatter over simulations divided by $\sqrt{N_{\rm sim}}$). The residuals are scattered around zero with random fluctuations of up to $\sim3\sigma$. {\sl Right:} exact bandpower window functions for the fully accounting for the effects of mode-coupling (solid black) compared with the input top-hat bandpower weights (red). Using the correct window functions is in general important in order to avoid parameter biases. Results are shown for the $BB$ power spectrum.} \label{fig:cmb_val_cls}
      \end{figure*}
      \begin{figure}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_sigmas_cmb_sph}
        \caption{1$\sigma$ errors on the $BB$, $EB$ and $EE$ power spectra (solid, dashed and dot-dashed lines respectively) for the standard pseudo-$C_\ell$ estimator (blue) and for the estimator with purified $B$ modes (red). $B$-mode purification is able to produce significant gains in sensitivity for the $B$-mode power spectrum, particularly at low $\ell$.}\label{fig:cmb_val_sigma}
      \end{figure}
      To quantify the presence of any residual bias in the estimator we start by computing, for each simulation and cross-power spectrum in the validation suite, the quantity:
      \begin{equation}
        \chi^2_i=({\bf s}_i-{\bf t})^T\,{\sf Cov}^{-1}\,({\bf s}_i-{\bf t}),
      \end{equation}
      where ${\bf s}_i$ is the vector of $N_{\rm dof}$ power spectrum values estimated in the $i$-th simulation, ${\bf t}$ is the analytical prediction for the input power spectrum, and ${\sf Cov}$ is the covariance matrix of the estimator, which we compute by averaging over all simulations:
      \begin{equation}
        {\sf Cov}=\frac{1}{N_{\rm sim}}\sum_{i=1}^{N_{\rm sim}}({\bf s}_i-\bar{\bf s})({\bf s}_i-\bar{\bf s})^T,\hspace{6pt} \bar{\bf s}\equiv\frac{1}{N_{\rm sim}}\sum_{i=1}^{N_{\rm sim}}{\bf s}_i.
      \end{equation}
      The full covariance matrix is very close to diagonal, as shown in the right panel of Fig. \ref{fig:lss_val_cls}. Under the assumption that the estimator is unbiased ($\langle{\bf s}\rangle={\bf t}$), the values of $\chi^2_i$ should follow a $\chi^2$ distribution with a $N_{\rm dof}$ degrees of freedom. Figure \ref{fig:lss_val_chi2} shows the distribution of $\chi^2_i$ across simulations (blue histograms), compared with the predicted $\chi^2$ distribution for $N_{\rm dof}=XXX$ (marked by the vertical dashed lines) for the six different power spectrum combinations. Both distributions are found to be compatible. Moreover, the dot-dashed lines show the values of $\chi^2$ found for the mean residual with respect to the input power spectra (lower panel of Fig. \ref{fig:lss_val_cls}). The probability to exceed (PTE) associated with these $\chi^2$ values are higher than $10\%$ in all cases, and therefore our validation suite does not show any statistically significant bias in the estimator. This bias, if any, must be at least $\sqrt{1000}\sim30$ times smaller than the statistical uncertainties associated with our validation suite, which are representative of the statistical power of next-generation galaxy surveys.

@@

    This paper presents \nmt, the first robust, validated and easy-to-use power spectrum estimation software package that implements the pseudo-$C_\ell$ method including all its generalizations (flat-sky, deprojection and purification). We have described the steps taken to validate the code, showing that it is able to provide unbiased results in a wide variety of scenarios. In particular, we have constructed a realistic validation suite that resembles the type of data, masks and contaminants that will affect future wide-area optical galaxy surveys and ground-based CMB polarization experiments. As shown in Section \ref{sec:validation}, we are able to recover the input power spectrum in all cases (which include varying degrees of contamination, purification, different spins etc.) with no detectable bias down to $\lesssim1/30$-th of the statistical uncertainties.

    Although \nmt~is intended to make power spectrum estimation as easy as possibe, some tasks are too dependent upon the properties of the data to be analyized to automatize them in a general way, and users are therefore in charge of them. We list the most important of these here:
    \begin{itemize}
      \item \nmt~operates directly on {\sl maps}. Galaxy survey data, however, are normally provided in the form of object catalogs, and the conversion those into the desired maps (e.g. galaxy overdensity or shear) must be done externally. 
      \item $E/B$ purification can only be carried out in the context of pseudo-$C_\ell$ estimators when the mask is differentiable up to its second derivatives, which can usually be achieved with an appropriate apodization. Although \nmt~provides tools to carry our this type of operations, some of the properties of the input mask (e.g. sharp corners) will in general spoil differentiability even after apodization. Users must therefore take care to inspect the mask derivatives before making use of \nmt's purification feature.
      \item A crucial part of any quadratic power spectrum estimation is the removal of noise bias in the auto-correlation of a given dataset. Although we have provided analytic formulas to estimate this bias in certain simple scenarios (see Section \ref{ssec:maths.beam_noise}), these are often not applicable. Users must therefore make sure that a sufficiently accurate estimate of the noise bias can be achieved (e.g. through simulations) or to use only cross-correlation data.
      \item Contaminant deprojection will be able to provide an accurate estimate of the power spectrum as long as two conditions are met. First, users must be able to provide a comprehensive list of possible contaminants as would be observed in their data (e.g. beam-smoothed, mean-subtracted etc.). Second, a good estimate of the underlying signal power spectra must be provided in order to calculate and subtract the deprojection bias. We have outlined a possible iterative process to do so in Section \ref{ssec:maths.deproj}.
    \end{itemize}

@@

\section{Full-sky validation for different pixel scales}
\label{app:full_sky_nside}
We checked that \texttt{NaMaster} returns unbiased power-spectra for all possible values of $N_{side}$. We show here some examples for 100 realisations of $N_{side}=128$ and $N_{side}=4096$.

\bsp

>>>>

    The paper is structured as follows: Section \ref{sec:maths} introduces the pseudo-$C_\ell$ estimator in the curved sky, including mode deprojection and $E/B$ purification. This is then generalized in the flat-sky approximation. In Section \ref{sec:structure} we briefly describe the implementation of these methods on \nmt, as well as its structure and usage. Section \ref{sec:validation} presents the stress tests used to validate the code. We envisage this code to be useful in analysis of both the Cosmic Microwave Background data and in the tomographic large-scale structure from photometric data and this section illustrates not just its validity but also applicability in typical user-case scenarios in both fields. We conclude in Section \ref{sec:discussion}. The appendices provide further details about the mathematical background used throughout the paper, and present some additional validation tests.

@@

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \va^\dagger \mC^{-1}
          \left(\frac{\partial \mC}{\partial B_i} \right)
          \mC^{-1} \va - \rm{Tr}\left[\left(\frac{\partial \mC}{\partial B_i}
          \right) {\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix,the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}
          \mC = \sum_i \left(\frac{\partial \mC}{\partial B_i}\right) + {\sf N},
        \end{equation}
        and the last term correspond to bias of convolved estimates by the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices $\partial {\sf C}/\partial B_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

      Two further points must be addressed before moving on. First, the computation of the deprojection bias requires an estimate of the true power spectrum ${\sf C}^{ab}_\ell$ (see Eqs. \ref{eq:dp.g.aux} and \ref{eq:dp.f.aux}). In practice, this can be done through an iterative approach, starting from an initial guess of ${\sf C}^{ab}_\ell$ and replacing it by the estimated power spectrum in the previous iteration. A sufficiently good guess will likely be enough to remove the bias in a single iteration. Alternatively, the ensemble average of Eq. \ref{eq:dpbias0} can be seen as a convolution of ${\sf C}^{ab}_\ell$, which could be treated by computing the convolution matrix as a modified version of the standard mode-coupling matrix that accounts for mode deprojection. Estimating this convolution matrix in this case becomes significantly more computationally intensive, and therefore an iterative approach is preferred. \as{I think this is not quite true. This latter way of looking at it, in effect replaces the true power spectrum with its observed quantity, which is fine in practice, but conceptually not quite right, for the same reason e.g. error on power spectrum should be based on real rather than measured power as illustrated by low CMB quadrupole. I would just remove text after ``Alternatively...''} \dam{we can do this, but I'd like to understand what the problem with this is. Since you can understand a mask as a set of deprojected contaminants (one per masked pixel), the pseudo-$C_\ell$ mode-coupling matrix is basically a flavour of the equations above for one particular set of templates. In this sense then writing the equation above as a matrix-vector product and inverting that matrix (after binning etc.) seems like the pseudo-$C_\ell$ way to go. Are you saying that you'll get more optimal results by putting in a guess for the power spectrum because that's closer to what a QML estimator would do?}

@@

    The code is written in C and wrapped into {\tt python} to facilitate its use and its combination with other software libraries for astronomy. \nmt~itself is based on a number of these libraries:
    \begin{itemize}
      \item {\tt HEALPix} \citep{2005ApJ...622..759G} is currently the only pixellation scheme supported for full-sky calculations. Through {\tt HEALPix}, \nmt~also makes use of {\tt cfitsio}~\citep{1999ASPC..172..487P}.
      \item The basic full-sky operation carried out by \nmt~is the spherical harmonic transform. For this, the code makes use of {\tt libsharp}~\citep{2013A&A...554A.112R}.
      \item In flat sky, SHTs are replaced by 2D DFTs, for which \nmt~uses the {\tt FFTW} library~\citep{FFTW05}.
    \end{itemize}

@@

    \paragraph*{NmtField.} This class stores all the necessary information about one individual observed spin-0 or spin-2 field. A {\tt NmtField} object is defined by a sky mask or weights map (e.g. $v$ in Eq. \ref{eq:aobs}), 1 or 2 sky maps (depending on the spin), corresponding to an observation of the field on the sky (${\bf a}^{\rm obs}$ in Eq. \ref{eq:aobs}), a set of contaminant templates (${\bf f}_i$ in Eq. \ref{eq:aobs}), and the choice to purify or not the $E$ or $B$ mode component of the field. Once initialized, this class carries out the following operations:
    \begin{itemize}
      \item SHT or DFT of the field (Eqs. \ref{eq:sht1} or \ref{eq:dft}).
      \item Contaminant deprojection (Eq. \ref{eq:clean}).
      \item $E$-mode and/or $B$-mode purification (Eq. \ref{eq:pureb}).
    \end{itemize}

@@

        Large-scale structure data are often characterized by inhomogeneous sky coverages and complex mask structures, associated with the presence of bright stars and observational artifacts. They are also affected by many sky contaminants, associated to Galactic sources (dust absorption, star density) and observing conditions (seeing, airmass etc.)~\citet{2016ApJS..226...24L}. The typically flatter power spectra for both galaxy clustering and lensing (e.g. when compared with the damping tail of CMB temperature fluctuations), makes the pseudo-$C_\ell$ approach ideal for this kind of analysis~\citet{2013MNRAS.435.1857L,2016ApJS..226...24L}. This science case therefore allows us to validate several aspects of the code simultaneously:
        \begin{itemize}
          \item The unbiased reconstruction of the 6 possible correlations between spin-0 and spin-2 fields.
          \item The inclusion of inhomogeneous noise and the correction of the associated noise bias.
          \item The effects of deprojection for several contaminant templates in different-spin fields.
          \item The code's ability to handle complex masks and high-resolution pixellation.
        \end{itemize}

@@

    \subsection{Full-sky large-scale structure validation}\label{ssec:validation.lss_full}
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_lss_sph}
        \includegraphics[width=0.99\columnwidth]{val_covar_lss_sph}
        \caption{{\sl Left panel:} average over all simulations in our full-sky LSS validation suite of the six power spectra between $\delta$, $\gamma_E$ and $\gamma_B$ (solid lines). The analytical prediction for the input signal power spectra are shown as dashed lines, and lie mostly below the solid lines (for the non-zero cases $\delta\times\delta$, $\delta\times\gamma_E$ and $\gamma_E\times\gamma_E$). The null power spectra ($\delta\times\gamma_B$, $\gamma_E\times\gamma_B$ and $\gamma_B\times\gamma_B$) are compatible with zero. The bottom subplot shows the mean residuals with respect to the input power spectra normalized by the 1$\sigma$ error over the mean (i.e. the scatter over simulations divided by $\sqrt{N_{\rm sim}}$). The residuals are scattered around zero and their fluctuations are well within those expected under the assumption of Gaussian statistics given 1000 independent realizations (see Fig. \ref{fig:lss_val_chi2}). {\sl Right panel:} correlation matrix ($C_{ij}/\sqrt{C_{ii}C_{jj}}$) of all measured power spectra estimated from 1000 realizations.} \label{fig:lss_val_cls}
      \end{figure*}
      \begin{figure*}
        \centering
        \includegraphics[width=0.80\textwidth]{val_chi2_lss_sph}
        \caption{Distribution of the $\chi^2$ values found for each of the 1000 simulations in our full-sky LSS validation suite (blue histograms). Results are shown for all possible cross-correlations, labeled in the top right corner of each subplot. The simulated distributions are compared with the expected $\chi^2$ distribution (solid lines) assuming Gaussian statistics for $N_{\rm dof}=681$ degrees of freedom (denoted by the dashed vertical lines). The dot-dashed vertical lines mark the $\chi^2$ value found for the mean residual with respect to the input power spectrum (see lower half of the left panel in Fig. \ref{fig:lss_val_cls}). The associated PTE values are $\gtrsim0.1$, and thus no statistically significant bias is found by our validation suite.} \label{fig:lss_val_chi2}
      \end{figure*}
      We generate 1000 realizations of contaminated maps of the galaxy overdensity $\delta$ and cosmic shear $\gamma$ using the models described in Section \ref{sssec:validation.suite.lss} with resolution $N_{\rm side}=1024$. In our fiducial run, we included all contaminants, and used a non-binary mask that traces the shot-noise fluctuations due to the presence of stars and depth variations. For each simulation, we use \nmt~to compute all six possible cross-correlations between the $\delta$ and the $E$ and $B$ modes of $\gamma$ (which we will label $\gamma_{E.B}$), including map-level contaminant deprojection and removal of the noise and deprojection bias in the power spectrum. We use band-powers defined by averaging over groups of $\Delta\ell=3$ multipoles.

@@

      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_cmb_sph}
        \includegraphics[width=0.99\columnwidth]{val_cmb_bpw}
        \caption{{\sl Left:} average over all simulations in our full-sky CMB validation suite of the three spin-2 power spectra ($EE$, $EB$ and $BB$, solid lines). The analytical prediction for the input signal power spectra are shown as dashed lines, and line mostly below the solid lines for the non-zero cases ($EE$ and $BB$). The null power spectrum $EB$ is compatible with zero. For comparison, the red dot-dashed line shows the result of not correcting for the mode-deprojection bias when estimating the $B$-mode power spectrum, which leads to a strong bias on small $\ell$. The bottom panel of this figure shows the mean residuals with respect to the input power spectra normalized by the $1\sigma$ error on the mean (i.e. the scatter over simulations divided by $\sqrt{N_{\rm sim}}$). The residuals are scattered around zero with random fluctuations of up to $\sim3\sigma$. {\sl Right:} exact bandpower window functions for the fully accounting for the effects of mode-coupling (solid black) compared with the input top-hat bandpower weights (red). Using the correct window functions is in general important in order to avoid parameter biases. Results are shown for the $BB$ power spectrum.} \label{fig:cmb_val_cls}
      \end{figure*}
      \begin{figure}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_sigmas_cmb_sph}
        \caption{1$\sigma$ errors on the $BB$, $EB$ and $EE$ power spectra (solid, dashed and dot-dashed lines respectively) for the standard pseudo-$C_\ell$ estimator (blue) and for the estimator with purified $B$ modes (red). $B$-mode purification is able to produce significant gains in sensitivity for the $B$-mode power spectrum, particularly at low $\ell$.}\label{fig:cmb_val_sigma}
      \end{figure}
      To quantify the presence of any residual bias in the estimator we start by computing, for each simulation and cross-power spectrum in the validation suite, the quantity:
      \begin{equation}
        \chi^2_i=({\bf s}_i-{\bf t})^T\,{\sf Cov}^{-1}\,({\bf s}_i-{\bf t}),
      \end{equation}
      where ${\bf s}_i$ is the vector of $N_{\rm dof}$ power spectrum values estimated in the $i$-th simulation, ${\bf t}$ is the analytical prediction for the input power spectrum, and ${\sf Cov}$ is the covariance matrix of the estimator, which we compute by averaging over all simulations:
      \begin{equation}
        {\sf Cov}=\frac{1}{N_{\rm sim}}\sum_{i=1}^{N_{\rm sim}}({\bf s}_i-\bar{\bf s})({\bf s}_i-\bar{\bf s})^T,\hspace{6pt} \bar{\bf s}\equiv\frac{1}{N_{\rm sim}}\sum_{i=1}^{N_{\rm sim}}{\bf s}_i.
      \end{equation}
      The full covariance matrix is very close to diagonal, as shown in the right panel of Fig. \ref{fig:lss_val_cls}. Under the assumption that the estimator is unbiased ($\langle{\bf s}\rangle={\bf t}$), the values of $\chi^2_i$ should follow a $\chi^2$ distribution with a $N_{\rm dof}$ degrees of freedom. Figure \ref{fig:lss_val_chi2} shows the distribution of $\chi^2_i$ across simulations (blue histograms), compared with the predicted $\chi^2$ distribution for $N_{\rm dof}=681$ (marked by the vertical dashed lines) for the six different power spectrum combinations. Both distributions are found to be compatible. Moreover, the dot-dashed lines show the values of $\chi^2$ found for the mean residual with respect to the input power spectra (lower panel of Fig. \ref{fig:lss_val_cls}). The probability to exceed (PTE) associated with these $\chi^2$ values are higher than $10\%$ in all cases, and therefore our validation suite does not show any statistically significant bias in the estimator. This bias, if any, must be at least $\sqrt{1000}\sim30$ times smaller than the statistical uncertainties associated with our validation suite, which are representative of the statistical power of next-generation galaxy surveys.

@@

    This paper presents \nmt, the first robust, validated and easy-to-use power spectrum estimation software package that implements the pseudo-$C_\ell$ method to compute angular cross-power spectra for any pair of spin-0 or spin-2 fields, including all their generalisations (flat-sky, deprojection and purification). We have described the steps taken to validate the code, showing that it is able to provide unbiased results in a wide variety of scenarios. In particular, we have constructed a realistic validation suite that resembles the type of data, masks and contaminants that will affect future wide-area optical galaxy surveys and ground-based CMB polarization experiments. As shown in Section \ref{sec:validation}, we are able to recover the input power spectrum in all cases (which include varying degrees of contamination, purification, different spins etc.) with no detectable bias down to $\lesssim1/30$-th of the statistical uncertainties.

    Although \nmt~is intended to make power spectrum estimation as easy as possible, some tasks are too dependent upon the properties of the data to be analyzed to automatize them in a general way, and users are therefore in charge of them. We list the most important of these here:
    \begin{itemize}
      \item \nmt~operates directly on {\sl maps}. Galaxy survey data, however, are normally provided in the form of object catalogs, and the conversion those into the desired maps (e.g. galaxy overdensity or shear) must be done externally. 
      \item $E/B$ purification can only be carried out in the context of pseudo-$C_\ell$ estimators when the mask is differentiable up to its second derivatives, which can usually be achieved with an appropriate apodization. Although \nmt~provides tools to carry our this type of operations, some of the properties of the input mask (e.g. sharp corners) will in general spoil differentiability even after apodization. Users must therefore take care to inspect the mask derivatives before making use of \nmt's purification feature.
      \item A crucial part of any quadratic power spectrum estimation is the removal of noise bias in the auto-correlation of a given dataset. Although we have provided analytic formulas to estimate this bias in certain simple scenarios (see Section \ref{ssec:maths.beam_noise}), these are often not applicable. Users must therefore make sure that a sufficiently accurate estimate of the noise bias can be achieved (e.g. through simulations) or to use only cross-correlation data.
      \item Contaminant deprojection will be able to provide an accurate estimate of the power spectrum as long as two conditions are met. First, users must be able to provide a comprehensive list of possible contaminants as would be observed in their data (e.g. beam-smoothed, mean-subtracted etc.). Second, a good estimate of the underlying signal power spectra must be provided in order to calculate and subtract the deprojection bias. We have outlined a possible iterative process to do so in Section \ref{ssec:maths.deproj}.
    \end{itemize}

@@

\section{Full-sky validation for different pixel scales}
\label{app:full_sky_nside}
We checked that \texttt{NaMaster} returns unbiased power-spectra for all possible values of $N_{side}$. In this appendix we show that we recover the correct power spectra using the inputs and foregrounds from Section \ref{sssec:validation.suite.lss}. In particular we tested $N_{side}$ ranging from 8 to 8192. For the latter, the computational resources needed make very difficult to generate and analyze a large enough number of simulations to have a good statistical measurement. However, we generate 100 realisations of $N_{side}=4096$ and show the results in Figure~\ref{fig:validation.fullsky.lss.4096}. We can see that the power spectra is recovered without any noticeable biases in the expected range of validity $\ell \leq 2 N_{side}$.

\begin{figure}
\centering
\includegraphics[width=0.9\columnwidth]{val_cl_lss_sph_4096}
\caption{Input (continuous darker lines) and measured (broken semi-transparent lines) power spectra using maps with \texttt{HEALPix} $N_{side}=4096$ using the inputs and foregrounds presented in Section \ref{sssec:validation.suite.lss}.}
\label{fig:validation.fullsky.lss.4096}
\end{figure}
\bsp
